\bf Supervised Tensor Decomposition with Multiple Side Information

### 2020-09-05 tensor_regress/jasa-latexsource/tensor-jasa.tex
revise
update

\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{algorithm}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor={blue},citecolor={blue},urlcolor={red}}

@@

\if1\blind
{
  \title{\bf Supervised Tensor Decomposition with Side Information on Multiple Modes}
  \author{Jiaxin Hu, Zhuoyan Xu, and Miaoyan Wang\\
    Department of Statistics, University of Wisconsin-Madison}
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Higher-order tensors have recently received increasing attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific applications often generate side information, including node features and interactions thereof, together with the tensor data. 
Identifying the relationship between a high-dimensional tensor and multi-mode features is an important yet challenging problem in neuroimaging, network analysis, and spatial-temporal modeling. 
Here, we develop a regression approach that incorporates side information as structured covariates, and present several case studies to show the benefit of proposed approach.  
Unlike unsupervised tensor decomposition, our supervised tensor factors capture the most effective dimension reduction of the data tensor confined to the feature spaces. 
%supervised tensor decomposition method that incorporates side information on multiple modes. 
%We consider the problem of decomposing a higher-order tensor with exponential-family entries, given side information on multiple modes. 
%Such data problems arise frequently in applications such as neuroimaging, network analysis, and spatial-temporal modeling. 
%We develop a regression approach that incorporates side information as structured covariates along multiple sides. 
%Unlike unsupervised tensor decomposition, the supervised approach finds the optimal low-rank factor matrices confined to the feature spaces. 
An efficient alternating updating algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. Our approach identifies the key global connectivity pattern and pinpoints the local regions that are associated with covariates.

@@

\noindent%
{\it Keywords:} Applications and case studies, Supervised dimension reduction, Exponential family observation, Reduced-rank tensor regression
\vfill

@@

Multi-dimensional arrays, known as tensors, are collected with additional covariates on multiple modes in many modern scientific and engineering studies. A popular example is the filed of neuroimaging~\citep{sun2017store,zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in the field of network analysis~\citep{baldin2018optimal,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent the friendships. Covariates such as people’s demographic information and the closeness of a friendship are often available. In both examples, it is of keen scientific interest to identify the variation in the array data (e.g., brain connectivity, social community patterns) that is affected by additional covariates. These seemingly different problems can be formulated as a regression problem in the tensor sense.

Emerging instances of the non-Gaussian datasets also require the effective tools applicable to a wide range of data, such as continuous, binary and count data. Examples include the Political Social science data \citep{hu2015scalable} which is a count dataset, and the brain connectivity network data \citep{wang2019common} which is usually encoded as binary datasets. Easily treating these datasets as Gaussian-valued data will yield unstable estimations and lead the predictions fall outside the valid range. A number of previous methods have been proposed \citep{yu2016learning, llosa2018tensor, raskutti2015convex, zhao2012higher} to address the tensor regression problem in various forms (e.g.\ scalar-to-tensor regression, tensor-response regression) from distinct perspectives. However, these methods assume the Gaussianity in the model, treating the input as Gaussian valued entries. Novel methods for exponential family tensor regression are necessary this time.

@@

{\bf Related work.} Our work is closely related to but also clearly distinctive from several lines of previous work.

The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor, hong2020generalized} that aims to find the best low-rank representation of data tensor. In contrast, our model can be viewed a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and covariates. The low-rank structure is determined jointly by the tensor response and matrix covariates.

The second line of work studies the scalar-to-tensor regression models, in which the response is a scalar and the \textit{predictor} is a tensor~\citep{zhou2013tensor,chen2019non}. Our proposal is orthogonal to theirs because we treat the tensor as a \textit{response}. The distinct model settings yield distinct focuses and interpretations. Take neuroimaging analysis as an example. The scalar-to-tensor regression aims to study the changes of the prediction outcomes when the brain connectivity network varies, while tensor-response model focuses on understanding the variation in the brain connectivity that may attribute to the individual features such as age, gender and disease status. Since scalar-to-tensor models do not utilize all the information in the response, they are not optimal when there exist correlations among response variables. In contrast, our model does not assume the independence among response entries, which allow the model to capture more structural information.

The third line of work studies the tensor-on-tensor regression models, in which both the response and predictor are tensors~\citep{raskutti2015convex, lock2018tensor, llosa2018tensor, gahrooei2020multiple}. Though our model actually tackles the similar problem as tensor-on-tensor regression, our work differs with previous work in several ways. First, previous methods mainly develop the model with Gaussian error. Their models are not suitable for general cases. Second, our study differs from previous work in the scope of results. \citep{gahrooei2020multiple} verifies the performance of estimation mainly via simulations, while our study provides a theoretical evidence on estimation. \citep{lock2018tensor, llosa2018tensor} perform solid inferences on parameter estimation, however, they do not assess the estimation and prediction errors, which are studied in our paper. Risk bounds are well studied in~\citep{raskutti2015convex} and our convergence rate coincides with their results in the order-3 low-rank case. Whereas, the convexity assumption of the regularization in their model yields a convex optimization problem. In contrast,  our model tackles a non-convex optimization problem, and employee distinct techniques.

{\bf Our contribution.} In this article, we present a general model and the associated theory for exponential family tensor regression. Exponential family tensor observation serves as the response, and the node features and/or the interactions form the matrix predictor. Figure~\ref{fig:intro1}b illustrates the general set-up we consider. Due to the generalized framework, our models allows heteroscedasticity by relating the variance of tensor entry to its mean. This ﬂexibility is particularly important in practice.  The regression approach allows the identiﬁcation of variation in the data tensor that is explained by the covariates. We utilize a low-rank constraint in the regression coeﬃcient to encourage the sharing among tensor entries. The statistical convergence of our estimator is established, and we quantify the gain in predictive power.

@@

We begin by reviewing the basic properties about tensors~\citep{kolda2009tensor}. We use $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ to denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as
\begin{equation}
\tY \times_1\mX_1\ldots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\ldots x_{j_K,i_K}^{(K)}},
\end{equation}
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use shorthand notion $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as $\langle \tY, \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}$. The Frobenius norm of tensor $\tY$ is defined as $\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}$. A higher-order tensor can be reshaped into a lower-order object~\citep{wang2017operator}. We use $\text{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ the operation that reshapes the tensor along mode-$k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The Tucker rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\text{Unfold}_k(\tY)$, $k=1,\ldots,K$. We use lower-case letters (e.g.,\ $a,b,c$) for scalars/vectors, upper-case boldface letters (e.g.,\ $\mA,\mB,\mC,$) for matrices, and calligraphy letters (e.g.,\ $\tA, \tB, \tC,$) for tensors of order three or greater. We let $\mI_d$ denote the $d \times d$ identity matrix, $[d]$ denote the $d$-set $\{1,\ldots,d\}$, and allow an $\mathbb{R}\to \mathbb{R}$ function to be applied to tensors in an element-wise manner.

\section{Motivation and model}\label{sec:model}
Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose we observe covariates on some of the $K$ modes. Let $\mX_k\in\mathbb{R}^{d_k\times p_k}$ denote the available covariates on the mode $k$, where $p_k\leq d_k$. We propose a multilinear structure on the conditional expectation of the tensor. Specifically, 
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\ \text{with}\\
&\Theta =\tB\times\{\mX_1,\ldots,\mX_K\} ,
\end{align}
where  $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the parameter tensor of interest, $f(\cdot)$ applied to $\Theta$ entrywise is a known link function in generalized linear model, and $\times$ denotes the tensor Tucker product. The choice of link function depends on the distribution of the response data. Some common choices are identity link for Gaussian tensor, logistic link for binary tensor, and $\exp(\cdot)$ link for Poisson tensor (see Table~\ref{table:link}).

@@

In our context, we model the the entries in the response tensor $y_{ijk}$ conditional on $\theta_{ijk}$ as independent draws from an exponential family. The quasi log-likelihood of~\eqref{eq:tensormodel2} is equal (ignoring constant) to Bregman distance between $\tY$ and $b'(\Theta)$:
\begin{align}
\tL_{\tY}(\tB)&=\langle \tY, \Theta \rangle - \sum_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K}),\\
\text{where}\ \Theta&=\tB\times\{\mX_1,\ldots,\mX_K\}.
\end{align}
We assume that we have an additional information on an upper bound $\alpha>0$ such that $\mnormSize{}{\Theta}\leq \alpha$. This is the case for many applications we have in mind such as brain network analysis where fiber connections are bounded. We propose a constrained maximum likelihood estimator (MLE) for the coefficient tensor:
\begin{align} \label{eq:MLE} 
	\hat \tB &=\argmax_{\hat \tB \colon \text{rank}(\tB)\leq \mr, \mnormSize{}{\Theta(\tB)}\leq \alpha  } \tL_{\tY}(\tB).
\end{align}
In the following theoretical analysis, we assume the rank $\mr=(r_1,\ldots,r_K)$ is known and fixed. The adaptation of unknown $\mr$ will be addressed in Section~\ref{sec:tuning}.

@@

\begin{cor}[Spatio-temporal growth model] The estimated type-by-time-by-space coefficient tensor converges at the rate $\tO\left({p+r+s\over dmn}\right)$ where $p\leq d$, $r\leq m$ and $s\leq n$. The estimation achieves consistency as long as the dimension grows in either of the three modes.
\end{cor}

\begin{cor} [Network population model] The estimated node-by-node-by-covariate tensor converges at the rate $\tO\left({2d+p\over d^2n}\right)$ where $p\leq n$. The estimation achieves consistency as the number of individuals or the number of nodes grows. 
\end{cor}

\begin{cor} [Dyadic data with node attributes] The estimated covariate-by-covariate matrix converges at the rate $\tO\left({p\over d^2}\right)$ where $p\leq d$. Again, our estimation achieves consistency as the number of nodes grows. 
\end{cor}

@@

@@

\section{Data analysis}
We apply our tensor regression model to two real datasets. The first application concerns the brain network modeling in response to individual attributes (i.e.\ covariate on one mode), and the second application focuses on multi-relational network analysis with dyadic attributes (i.e.\ covariates on two modes). 
\subsection{Human Connectome Project (HCP)}
The Human connectome project (HCP~\citep{HCP}) aims to build a ``network map'' that characterizes the anatomical and functional connectivity within healthy human brains. We take a subset of HCP data that consists of 136 brain structural networks, one for each individual. Each brain network is represented as a 68-by-68 binary matrix, where the entries encode the presence or absence of fiber connections between the 68 brain regions. We consider four individual-covariates: gender (65 females vs.\ 71 males), age 22-25 ($n=35$), age 26-30 ($n=58$), and age 31+ ($n=43$). The goal is to identify the connection edges that are affected by the individual covariates. A key challenge in brain network is that the edges are correlated; for example, two edges may stem out from a same brain region, and it is of importance to take into account the within-dyad dependence.

@@

\subsection{Nations data}

@@

@@

\begin{figure}[t]
\centering
\includegraphics[width=10cm]{coef.pdf}
\caption{Effect estimation in the \emph{Nations} data. Panels (a)-(d) represent the estimated effects of country-level attributes towards the connection probability, for relations \emph{warnning}, \emph{economicaid},\emph{intergovorg}, and \emph{commonblock}, respectively. }\label{fig:est}

@@

\begin{table}[H]
\begin{tabular}{c|l}
\hline
\multirow{4}{*}{Cluster I}  &officialvisits, intergovorgs,  militaryactions, violentactions, duration, \\
&negativebehavior, boycottembargo, aidenemy, negativecomm, accusation, \\
& protestsunoffialacts,    nonviolentbehavior, emigrants, relexports,   \\
& timesincewar, commonbloc2, rintergovorgs3, relintergovorgs\\
\hline
\multirow{4}{*}{Cluster II}& economicaid, booktranslations, tourism, relbooktranslations, releconomicaid, \\
& conferences, severdiplomatic, expeldiplomats, attackembassy, unweightedunvote,\\
& reltourism, tourism3, relemigrants, emigrants3, students, relstudents, \\
&exports, exports3, lostterritory, dependent, militaryalliance, warning  \\
\hline
\multirow{2}{*}{Cluster III} & treaties, reltreaties, exportbooks, relexportbooks, weightedunvote, ngo, \\
&relngo, ngoorgs3, embassy, reldiplomacy, timesinceally, independence, commonbloc1\\
\hline
Cluster IV & commonbloc0, blockpositionindex\\
\end{tabular}
\caption{$K$-means clustering of relations based on factor matrix in the coefficient tensor.}\label{tab:s1}
\end{table}

@@

\section{Additional results for real data analysis}

\subsection{HCP data analysis}

\begin{figure}[t]
\includegraphics[width=17cm]{compare.pdf}
\caption{Comparison of coefficient estimation in the HCP data.}\label{fig:s1}
\end{figure}

>>>>

\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{algorithm}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue}}

@@

\if1\blind
{
  \title{\bf Supervised Tensor Decomposition with Multiple Side Information}
  \author{Jiaxin Hu, Zhuoyan Xu, and Miaoyan Wang\\
    Department of Statistics, University of Wisconsin-Madison}
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. Identifying the relationship between a high-dimensional tensor and side information is important yet challenging. Here, we develop a tensor decomposition method that incorporates multiple side information as structured covariates, and present case studies to show the benefit of proposed approach. Unlike unsupervised tensor decomposition, our supervised decomposition captures the most effective dimension reduction of the data tensor confined to the feature space on each mode. 
%supervised tensor decomposition method that incorporates side information on multiple modes. 
%We consider the problem of decomposing a higher-order tensor with exponential-family entries, given side information on multiple modes. 
%Such data problems arise frequently in applications such as neuroimaging, network analysis, and spatial-temporal modeling. 
%We develop a regression approach that incorporates side information as structured covariates along multiple sides. 
%Unlike unsupervised tensor decomposition, the supervised approach finds the optimal low-rank factor matrices confined to the feature spaces. 
An efficient alternating updating algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We identify the key global connectivity pattern and pinpoints the local regions that are associated with covariates. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, the package and data used are available at ~\url{https://CRAN.R-project.org/package=tensorregress}.

@@

\noindent%
{\it Keywords:} Applications and case studies, Supervised dimension reduction, Exponential family observation, Tensor decomposition
\vfill

@@

Multi-dimensional arrays, known as tensors, are often collected with side information on multiple modes in modern scientific and engineering studies. A popular example is in neuroimaging~\citep{sun2017store,zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in the field of network analysis~\citep{baldin2018optimal,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent the friendships. Side information such as people’s demographic information and the closeness of a friendship are often available. In both examples, it is of keen scientific interest to identify the variation in the tensor observation (e.g., brain connectivities, social community patterns) that is affected by available feature data. These seemingly different scenarios pose a common yet challenging problem for tensor modeling.

In addition to the aforementioned challenges, many instances of tensor observation consist of non-Gaussian measurements. Examples include the political interaction dataset \citep{hu2015scalable} which measures action counts between countries for various relationship, and the brain connectivity network data \citep{wang2019common} which is encoded as binary adjacency matrices. Classical tensor decomposition methods are based on minimizing the Frobenious norm of the reconstruction error, leading to suboptimal predictions for binary or count responses. A number of supervised tensor methods have been proposed \citep{yu2016learning, raskutti2015convex, zhao2012higher} to address the tensor regression problem in various forms (e.g.\ scalar-to-tensor regression, tensor-response regression). These methods often assume Gaussian distribution for the tensor entries, or impose random designs for the covariate matrices, both of which are less suitable for applications of our interest. The gap in theory and practice suggests a huge opportunity for improved modeling paradigms that better capture the complexity in tensor data with side information.

@@

We present a modeling paradigm and associated method for analyzing exponential family tensors with multiple side information. 
We formulate the learning task as a structured regression problem, with tensor observation serving as the response, and the side information forming multiple covariate matrices. Figure~\ref{fig:intro1}b illustrates our model in the special case of order-3 tensors. A low-rank structure is imposed to the mean tensor $\mathbb{E}(\tY)\in\mathbb{R}^{d_1\times d_2\times d_3}$, where unlike classical decomposition, the tensor factors are assumed to belong to the subspace spanned by known features $\mX_k\in\mathbb{R}^{p_k\times d_k}$ for $k=1,2,3$. The unknown matrix $\mM_k\in\mathbb{R}^{p_k\times r_k}$ link the data tensor to each of the feature matrices, thus allowing the identification the variation in the tensor entries attributable to the side information

In addition, we use ideas from generalized linear model (GLM) to allow heteroscedacity due to mean-variance relationship across tensor entries. This flexibility is important in practice. In classical GLM, the sample size $n$ and feature dimension $d$ are well defined; however, in the tensor data analysis, we observe only one single realization of tensor response but up to $K$ feature matrices, where $K$ is the number of tensor modes.  Both the sample size and feature dimension grow exponentially in $K$. We overcome the high-dimensionality using a possibly non-linear, low-rank constraint in the mean tensor, thereby encouraging the sharing among entries in the data tensor. The statistical convergence of our estimator is established, and we quantify the gain in predictive power through several case studies.

Our work is closely related to but also clearly distinctive from several lines of previous work. The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor, hong2020generalized} that aims to find the best low-rank representation of data tensor. In contrast, our model can be viewed a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and covariates. The low-rank structure is determined jointly by the tensor response and matrix covariates.

The second line of work studies the scalar-to-tensor regression models, in which the response is a scalar and the \textit{predictor} is a tensor~\citep{zhou2013tensor,chen2019non}. Our proposal is orthogonal to theirs because we treat the tensor as random variables. The distinct model settings yield distinct focuses and interpretations. Take neuroimaging analysis as an example. The scalar-to-tensor regression aims to study the changes of the disease status when the brain connectivity network varies, while tensor-response model focuses on understanding the variation in the brain connectivity that is attribute to the individual features such as age, gender and disease status. Since scalar-to-tensor models do not model the stochastic noise in the tensor, they are not suited for dimension reduction for tensor data.

The third line of work studies the tensor-on-tensor regression models, in which both the response and predictor are tensors~\citep{raskutti2015convex, lock2018tensor, gahrooei2020multiple}.

Though our model actually tackles the similar problem as tensor-on-tensor regression, our work differs with previous work in several ways. Previous methods mainly develop the model with Gaussian error. Their models are not suitable for general cases. Second, our study differs from previous work in the scope of results. \citep{gahrooei2020multiple} verifies the performance of estimation mainly via simulations, while our study provides a theoretical evidence on estimation. \citep{lock2018tensor} perform Bayesian inferences on parameter estimation, however, they do not assess the estimation and prediction errors, which are studied in our paper. Risk bounds are well studied in~\citep{raskutti2015convex} and our convergence rate coincides with their results in the order-3 low-rank case. Whereas, the convexity assumption of the regularization in their model yields a convex optimization problem. In contrast,  our model tackles a non-convex optimization problem, and employee distinct techniques.

@@

We summarize the basic tensor properties used in the paper; more details can be found in~\citet{kolda2009tensor}. We use $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ to denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as
\begin{equation}
\tY \times_1\mX_1\ldots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\ldots x_{j_K,i_K}^{(K)}},
\end{equation}
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use shorthand notion $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as $\langle \tY, \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}$. The Frobenius norm of tensor $\tY$ is defined as $\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}$. A higher-order tensor can be reshaped into a lower-order object~\citep{wang2017operator}. We use $\text{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ the operation that reshapes the tensor along mode-$k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The Tucker rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\text{Unfold}_k(\tY)$, $k=1,\ldots,K$. We use lower-case letters (e.g.,\ $a,b,c$) for scalars/vectors, upper-case boldface letters (e.g.,\ $\mA,\mB,\mC,$) for matrices, and calligraphy letters (e.g.,\ $\tA, \tB, \tC,$) for tensors of order three or greater. We let $\mI_d$ denote the $d \times d$ identity matrix, $[d]$ denote the $d$-set $\{1,\ldots,d\}$, and allow an $\mathbb{R}\to \mathbb{R}$ function to be applied to tensors in an element-wise manner.

\section{Motivation and model}\label{sec:model}
We start with general model formulation and then provide concrete examples that motivate the model. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose we observe covariates on some of the $K$ modes. Let $\mX_k\in\mathbb{R}^{d_k\times p_k}$ denote the available covariates on the mode $k$, where $p_k\leq d_k$. We consider a multilinear structure on the conditional expectation of the tensor. Specifically, 
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\ \text{with}\\
&\Theta =\tB\times\{\mX_1,\ldots,\mX_K\} ,
\end{align}
where  $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the parameter tensor of interest, $f(\cdot)$ applied to $\Theta$ entrywise is a known link function in generalized linear model, and $\times$ denotes the tensor Tucker product. The choice of link function depends on the distribution of the response data. Some common choices are identity link for Gaussian tensor, logistic link for binary tensor, and $\exp(\cdot)$ link for Poisson tensor (see Table~\ref{table:link}).

@@

In our context, we model the the entries in the response tensor $y_{ijk}$ conditional on $\theta_{ijk}$ as independent draws from an exponential family. The quasi log-likelihood of~\eqref{eq:tensormodel2} is equal (ignoring constant) to Bregman distance between $\tY$ and $b'(\Theta)$:
\begin{align}
\tL_{\tY}(\tB)&=\langle \tY, \Theta \rangle - \sum_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K}),\\
\text{where}\ \Theta&=\tB\times\{\mX_1,\ldots,\mX_K\}.
\end{align}
We assume that we have an additional information on an upper bound $\alpha>0$ such that $\mnormSize{}{\Theta}\leq \alpha$. We propose a constrained maximum likelihood estimator (MLE) for the coefficient tensor:
\begin{align} \label{eq:MLE} 
	\hat \tB &=\argmax_{\hat \tB \colon \text{rank}(\tB)\leq \mr, \mnormSize{}{\Theta(\tB)}\leq \alpha  } \tL_{\tY}(\tB).
\end{align}
In the following theoretical analysis, we assume the rank $\mr=(r_1,\ldots,r_K)$ is known and fixed. The adaptation of unknown $\mr$ will be addressed in Section~\ref{sec:tuning}.

@@

\begin{example}[Spatio-temporal growth model] The estimated type-by-time-by-space coefficient tensor converges at the rate $\tO\left({p+r+s\over dmn}\right)$ where $p\leq d$, $r\leq m$ and $s\leq n$. The estimation achieves consistency as long as the dimension grows in either of the three modes.
\end{example}

\begin{example} [Network population model] The estimated node-by-node-by-covariate tensor converges at the rate $\tO\left({2d+p\over d^2n}\right)$ where $p\leq n$. The estimation achieves consistency as the number of individuals or the number of nodes grows. 
\end{example}

\begin{example} [Dyadic data with node attributes] The estimated covariate-by-covariate matrix converges at the rate $\tO\left({p\over d^2}\right)$ where $p\leq d$. Again, our estimation achieves consistency as the number of nodes grows. 
\end{example}

@@

\subsection{Missing values}
.....

@@

\section{Data analysis}
We apply our tensor regression model to two datasets. The first application concerns the brain network modeling in response to individual attributes (i.e.\ covariate on one mode), and the second application focuses on multi-relational network analysis with dyadic attributes (i.e.\ covariates on two modes).

\subsection{Analysis of human brain connection data}
The Human Connectome Project (HCP) aims to build a ``network map'' that characterizes the anatomical and functional connectivity within healthy human brains~\citep{HCP}. We take a subset of HCP data that consists of 136 brain structural networks, one for each individual. Each brain network is represented as a 68-by-68 binary matrix, where the entries encode the presence or absence of fiber connections between the 68 brain regions. We consider four individual-covariates: gender (65 females vs.\ 71 males), age 22-25 ($n=35$), age 26-30 ($n=58$), and age 31+ ($n=43$). The goal is to identify the connection edges that are affected by the individual covariates. A key challenge in brain network is that the edges are correlated; for example, two edges may stem out from a same brain region, and it is of importance to take into account the within-dyad dependence.

@@

\subsection{Analysis of political relational data}

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{coef.pdf}
\caption{Effect estimation in the \emph{Nations} data. Panels (a)-(d) represent the estimated effects of country-level attributes towards the connection probability, for relations \emph{warnning}, \emph{economicaid},\emph{intergovorg}, and \emph{commonblock}, respectively. }\label{fig:est}

@@

\begin{table}[H]
\begin{tabular}{c|c|l}
\hline
\multirow{4}{*}{Cluster I}  & Government &officialvisits, intergovorgs,  militaryactions, violentactions, duration, \\
&&negativebehavior, boycottembargo, aidenemy, negativecomm, accusation, \\
& &protestsunoffialacts,    nonviolentbehavior, emigrants, relexports,   \\
& &timesincewar, commonbloc2, rintergovorgs3, relintergovorgs\\
\hline
\multirow{4}{*}{Cluster II}& Economics &economicaid, booktranslations, tourism, relbooktranslations, releconomicaid, \\
&& conferences, severdiplomatic, expeldiplomats, attackembassy, unweightedunvote,\\
&& reltourism, tourism3, relemigrants, emigrants3, students, relstudents, \\
&&exports, exports3, lostterritory, dependent, militaryalliance, warning  \\
\hline
\multirow{2}{*}{Cluster III} & treaties &treaties, reltreaties, exportbooks, relexportbooks, weightedunvote, ngo, \\
&&relngo, ngoorgs3, embassy, reldiplomacy, timesinceally, independence, commonbloc1\\
\hline
Cluster IV & Politics &commonbloc0, blockpositionindex\\
\end{tabular}
\caption{$K$-means clustering of relations based on factor matrix in the coefficient tensor.}\label{tab:s1}
\end{table}

@@

@@

\begin{figure}[H]
\includegraphics[width=17cm]{compare.pdf}
\caption{Comparison of coefficient estimation in the HCP data.}\label{fig:s1}
\end{figure}
